---
title: "Assignment 3: Model comparison"
author: "Marton Kovacs/Zoltan Kekecs"
output: html_document
editor_options: 
  chunk_output_type: console
---

In this lab assignment you are going to work with (simulated) data related to perioperative pain and its psychological and hormonal predictors. In the assignment you will assess the added benefit of including some psychological and hormonal predictors to the already established demographic predictors of pain.

In this assignment you will set up a hierarchical regression model to predict postoperative pain after wisdom tooth surgery. 

# Research problem

The amount of pain experienced around and after surgeries are highly variable between and within individuals. In order to improve surgical pain management regimens we need to understand what influences pain around surgical procedures and predict the amount of pain an individual will experience.

Your first study in this area is related to assessing the influence of trait and state psychological measures on pain, and to see whether taking into account these variables can improve our understanding of postoperative pain.

# Procedures and measures

Use the data file called ‘assignment_3_dataset’, from the 'data/' folder.

You have collected data from 160 adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the serum (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset.

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgmental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, serum cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.

# Research question

Previous studies and meta-analyses showed that age and sex are often predictors of pain (age is negatively associated with pain, while sex is a predictor more dependent on the type of the procedure). You would like to determine the extent to which taking into account psychological and hormonal variables aside from the already used demographic variables would improve our understanding of postoperative pain.

To answer this research question you will __need to compare two models__ (with a hierarchical regression). The __simpler model__ should contain __age and sex as predictors of pain__, while the __more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__. Notice that the predictors used in the simpler model are a subset of the predictors used in more complex model. __You will have to do model comparison to assess whether substantial new information was gained about pain in the more complex model compared to the simpler model.__  

# What to report

As usual, before you can interpret your model, you will need to run data and model diagnostics. First, check the variables included in the more complex model (age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures as predictors, and pain as an outcome) for __coding errors__, and the model itself for __influential outliers__ (for example using Cook’s distance). Furthermore, check the final model to see if the __assumptions of linear regression hold true__, that is, __normality__ (of the residuals), __linearity__ (of the relationship), __homogeneity of variance__ (also called homoscedasticity) and that there is no excess __multicollinearity__ (“uncorrelated predictors” in Navarro’s words). If you find anything amiss during these checks, make the appropriate decision or correction and report your findings and actions in your report. 

__Note:__ If you do any changes, such as exclude cases, or exclude predictors from the model, you will have to re-run the above checks for your final data and model.

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

Write up the regression equation of the more complex model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile.

```{r}
setwd("~/Desktop/ELTE_PPK_PHD/Statisztikai Elemzes/Assignments")

#install.packages("readxl")
library(readxl)

# Read the Excel file
df <- read_excel("~/Desktop/ELTE_PPK_PHD/Statisztikai Elemzes/Assignments/assignment_3_dataset.xlsx")

# View the first few rows of the data frame
head(df)
```

## Data and model diagnostics 
### Data diagnostics
#### Descriptives of the variables

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r}
#Initial Exploration:
#Look at the structure of the dataset.
#View the first few rows.

library(tidyverse)

str(df)
head(df)


#Summary Statistics:
summary(df)

# For categorical columns:
table(df$sex)

#Visual Exploration
# Histogram for a numeric column, say "numeric_column":
ggplot(df, aes(x = pain)) + 
  geom_histogram(binwidth = 2, fill = "skyblue", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of pain")

ggplot(df, aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "red", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of age")

ggplot(df, aes(x = STAI_trait)) + 
  geom_histogram(binwidth = 3, fill = "yellow", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of STAI_trait")

ggplot(df, aes(x = pain_cat)) + 
  geom_histogram(binwidth = 3, fill = "green", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of pain_cat")

ggplot(df, aes(x = cortisol_serum)) + 
  geom_histogram(binwidth = 0.5, fill = "grey", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of cortisol_serum")

ggplot(df, aes(x = cortisol_saliva)) + 
  geom_histogram(binwidth = 0.75, fill = "white", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of cortisol_saliva")

ggplot(df, aes(x = mindfulness)) + 
  geom_histogram(binwidth = 0.5, fill = "orange", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of mindfulness")

ggplot(df, aes(x = weight)) + 
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of weight")

ggplot(df, aes(x = IQ)) + 
  geom_histogram(binwidth = 10, fill = "brown", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of IQ")

ggplot(df, aes(x = household_income)) + 
  geom_histogram(binwidth = 10000, fill = "violet", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of household_income")

# Bar plot for a categorical column:
ggplot(df, aes(x = sex)) + 
  geom_bar(fill = "skyblue", color = "black", alpha = 0.7) +
  labs(title = "Bar plot of Sex")

```

#### Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}
df$sex <- ifelse(df$sex == "woman", "female", df$sex)
```

### Model diagnostics
#### Build the more complex model

In order to test the more complex model for outliers and to test the assumptions first build the model.

```{r}
# Fit the multiple linear regression model
model <- lm(pain ~ age + STAI_trait + pain_cat + cortisol_serum + cortisol_saliva + mindfulness + weight + IQ + household_income + sex, data = df)

# Display a summary of the model
summary(model)

```

#### Checking for influential outliers

Check for outlier values in the model.

```{r}
# Find outliers and make a plot
cooksd <- cooks.distance(model)
plot(cooksd, pch = "*", cex = 2, main = "Cook's Distance")
abline(h = 4*mean(cooksd, na.rm=T), col="red")  # influential point threshold


# Find outliers and make a plot
std_residuals <- rstandard(model)
plot(std_residuals ~ fitted(model), ylab = "Standardized Residuals")
abline(h = c(-3, 3), col = "red")

# Find outliers
outliers <- which(abs(std_residuals) > 3)

# Outliers will now contain the row numbers of the data points that are outliers
print(outliers)

# Remove outliers from the dataset
data_cleaned <- df[-outliers, ]
```

```{r}
# Fit the multiple linear regression model again without the outlier
model.2 <- lm(pain ~ age + STAI_trait + pain_cat + cortisol_serum + cortisol_saliva + mindfulness + weight + IQ + household_income + sex, data = data_cleaned)

# Display a summary of the model
summary(model.2)
```


#### Checking assumptions

Check the normality assumption.

```{r}
qqnorm(residuals(model.2))
qqline(residuals(model.2))

shapiro.test(residuals(model.2))

ks.test(residuals(model.2), "pnorm", mean(residuals(model.2)), sd(residuals(model.2)))

hist(residuals(model.2), breaks=30, main="Histogram of Residuals", xlab="Residuals")
```

Check the linearity assumption.

```{r}
plot(model.2, which = 1)
```

Check the homoscedasticty assumption (homogeneity of variance).

```{r}
# Install and load the package
#install.packages("lmtest")
#A significant p-value indicates heteroscedasticity. 

library(lmtest)

bptest(model.2)
```

Check the multicollinearity assumption.

(VIF above 5), or a VIF threshold of 3 is recommended in this paper: http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x/full

Some info about VIF: 
https://statisticalhorizons.com/multicollinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

```{r}
# Install and load the package
#install.packages("car")

#A VIF value above 5-10 suggests high multicollinearity
library(car)

vif_values <- vif(model.2)
print(vif_values)
```

### Making decision based on model diagnostics

If based on the assumption tests you decide to drop a predictor variable you should do that here. Create your updated model.

```{r}
# Assuming you're dropping a predictor named 'drop_this_predictor'
#updated_model <- lm(response_variable ~ . - drop_this_predictor, data = df)
#summary(updated_model)

```


```{r}
# Calculate studentized residuals
stud_res <- rstudent(model.2)

# Typically, absolute values greater than 2 or 3 are considered large.
potential_outliers <- which(abs(stud_res) > 3)
print(potential_outliers)

```


```{r}
cook_d <- cooks.distance(model.2)

# A common rule of thumb is that observations with Cook's distance greater than 4/(n-k-1) might be outliers.
# where n is the number of observations and k is the number of predictors.
threshold <- 4/(length(cook_d) - length(coef(model.2)) - 1)
influential_obs <- which(cook_d > threshold)
print(influential_obs)

```

```{r}
# Let's say you've decided to remove based on Cook's distance
df_cleaned <- data_cleaned[-influential_obs, ]
```

#### Checking outliers of the updated model

```{r}
# Fit the multiple linear regression model
model.updated <- lm(pain ~ age + STAI_trait + pain_cat + cortisol_serum + cortisol_saliva + mindfulness + weight + IQ + household_income + sex, data = df_cleaned)

# Display a summary of the model
summary(model.updated)


cook_d <- cooks.distance(model.updated)

# A common rule of thumb is that observations with Cook's distance greater than 4/(n-k-1) might be outliers.
# where n is the number of observations and k is the number of predictors.
threshold <- 4/(length(cook_d) - length(coef(model.updated)) - 1)
influential_obs <- which(cook_d > threshold)
print(influential_obs)
```

#### Checking assumptions of the updated model

Normality assumption

```{r}
qqnorm(residuals(model.updated))
qqline(residuals(model.updated))

shapiro.test(residuals(model.updated))
```

Linearity assumption

```{r}
plot(model.updated, which = 1)
```

Homoscedasticty assumption (homogeneity of variance)

```{r}
bptest(model.updated)
```

Multicollinearity assumption

```{r}
#A VIF value above 5-10 suggests high multicollinearity
library(car)

vif_values <- vif(model.updated)
print(vif_values)
```

## Model comparison

Create the simple model and get the results of the model that needs to be reported based on the What to report section.

```{r}
#The __simpler model__ should contain __age and sex as predictors of pain_

# Fit the multiple linear regression model
model.simple <- lm(pain ~ age + sex, data = df_cleaned)

# Display a summary of the model
summary(model.simple)

```

Create the more complex model based on the results of the model diagnostics. Also, get the results that needs to be reported based on the What to report section.

```{r}
# Fit the multiple linear regression model
model.updated <- lm(pain ~ age + STAI_trait + pain_cat + cortisol_serum + cortisol_saliva + mindfulness + weight + IQ + household_income + sex, data = df_cleaned)

# Display a summary of the model
summary(model.updated)
```

Compare the two models.

```{r}
rsq_simple <- summary(model.simple)$r.squared
rsq_complex <- summary(model.updated)$r.squared

#R-squared measures the proportion of variability in the response variable that's explained by the model. A higher R-squared value typically indicates a better fit to the data, but it will almost always increase or stay the same as you add more predictors, even if they are irrelevant.
```

```{r}
adjrsq_simple <- summary(model.simple)$adj.r.squared
adjrsq_simple
adjrsq_complex <- summary(model.updated)$adj.r.squared
adjrsq_complex
```

```{r}
aic_simple <- AIC(model.simple)
aic_simple
aic_complex <- AIC(model.updated)
aic_complex

bic_simple <- BIC(model.simple)
bic_simple
bic_complex <- BIC(model.updated)
bic_complex

#Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) are measures used for model selection. Lower values of AIC or BIC are indicative of better-fitting models.
```

```{r}
#The anova() function can be used to compare the two models. If the complex model is significantly better than the simpler model, the p-value will be small.

anova(model.simple, model.updated)

```

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

```{r}
# Extract Adjusted R-squared from both models
adj_r2_simple <- summary(model.simple)$adj.r.squared
adj_r2_complex <- summary(model.updated)$adj.r.squared

# Display Adjusted R-squared
adj_r2_simple
adj_r2_complex

```

```{r}
# Extract F-statistic and p-value from both models
fstat_simple <- summary(model.simple)$fstatistic
fstat_complex <- summary(model.updated)$fstatistic

# Calculate p-values
pval_simple <- pf(fstat_simple[1], fstat_simple[2], fstat_simple[3], lower.tail = FALSE)
pval_complex <- pf(fstat_complex[1], fstat_complex[2], fstat_complex[3], lower.tail = FALSE)

# Display F-statistic and p-value
fstat_simple
pval_simple
fstat_complex
pval_complex

```
For the Simple Model: 
Adjusted R-squared: 0.1074
F-statistic: 9.902 on 2 and 146 degrees of freedom
p-value: 9.274e-05

```{r}
# Simple Model

# Load necessary libraries
library(knitr)
# Create a table of coefficients for the simple model
kable(summary(model.simple)$coefficients, caption = "Coefficients of the Simple Model")
```

For the Complex Model:

Adjusted R-squared: 0.4428
F-statistic: 12.76 on 10 and 138 degrees of freedom
p-value: 1.471e-15
```{r}
# Complex Model

# Create a table of coefficients for the complex model
kable(summary(model.updated)$coefficients, caption = "Coefficients of the Complex Model")
```

## Model Comparison
The Adjusted R-squared increased from 0.1074 in the simpler model to 0.4428 in the more complex model, indicating a substantial increase in the amount of variance in pain explained by including psychological and hormonal variables.

The F-statistic and its associated p-value also support the more complex model, indicating that the additional variables significantly improve the model.



## Regression Equation for the More Complex Model

The regression equation based on the estimated coefficients is as follows:

\[
\begin{aligned}
\text{Pain} = & \ 0.6561 \\
& - 0.02537 \times \text{Age} \\
& - 0.01710 \times \text{STAI\_trait} \\
& + 0.1156 \times \text{Pain\_Catastrophizing} \\
& - 0.05647 \times \text{Cortisol\_Serum} \\
& + 0.6668 \times \text{Cortisol\_Saliva} \\
& - 0.05839 \times \text{Mindfulness} \\
& - 0.01599 \times \text{Weight} \\
& + 0.004028 \times \text{IQ} \\
& + 0.000003276 \times \text{Household\_Income} \\
& + 0.3506 \times \text{Sex\_Male}
\end{aligned}
\]



Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

## Model Comparison Summary

Adjusted R-squared
The adjusted R-squared for the simpler model is 0.1074.
The adjusted R-squared for the more complex model is 0.4428.

This indicates that the more complex model explains approximately 44.28% of the variance in pain, which is a significant improvement over the 10.74% explained by the simpler model.

Akaike Information Criterion (AIC)

The AIC for the simpler model is 520.7289.
The AIC for the more complex model is 458.1112.

A lower AIC suggests a better model fit, given that it penalizes the complexity of the model. The substantial decrease in AIC for the more complex model suggests that it fits the data much better than the simpler model despite the additional parameters.

Bayesian Information Criterion (BIC)

The BIC for the simpler model is 532.7447.
The BIC for the more complex model is 494.1586.

Like AIC, a lower BIC value indicates a better model fit. The decrease in BIC when moving from the simpler model to the more complex model also suggests an improved fit, even after accounting for model complexity.

ANOVA Comparison (Likelihood Ratio Test)

The F-statistic for comparing the two models is 11.987.
The p-value for this test is 6.755e-13.

The ANOVA comparison shows that the more complex model provides a significantly better fit to the data than the simpler model (as indicated by the highly significant p-value).


```{r}
# Install and load required package
if (!require("kableExtra")) {
  install.packages("kableExtra")
}
library(kableExtra)

# Create comparison table for model fit statistics
comparison_table <- data.frame(
  Metric = c("Adjusted R-squared", "AIC", "BIC"),
  `Simpler Model` = c(0.1073817, 520.7289, 532.7447),
  `More Complex Model` = c(0.4428268, 458.1112, 494.1586)
)

# Create ANOVA comparison table
anova_results <- data.frame(
  `F-statistic` = 11.987,
  `p-value` = format.pval(6.755e-13, digits=2, eps=0.01)
)

# Generate tables with kable and kableExtra for better formatting
model_comparison_table <- kable(comparison_table, caption = "Model Fit Statistics Comparison", 
                                booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

anova_comparison_table <- kable(anova_results, caption = "ANOVA Comparison of the Two Models", 
                                booktabs = TRUE) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

# Print tables
model_comparison_table
anova_comparison_table

```


# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

The hierarchical regression analysis comparing the two models shows that the inclusion of psychological and hormonal measures significantly improves the ability of the model to explain the variance in postoperative pain. This is evidenced by the substantial increase in the Adjusted R-squared value from 0.1074 in the simpler model to 0.4428 in the more complex model, indicating that a much larger proportion of the variance in pain is accounted for when these additional variables are included.

The Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC) values are both lower for the more complex model (AIC: 458.1112, BIC: 494.1586) compared to the simpler model (AIC: 520.7289, BIC: 532.7447). Lower values of AIC and BIC suggest that the more complex model has a better fit to the data, despite the larger number of predictors.

The F-test from the ANOVA comparing the two models is highly significant (F = 11.987, p < 0.001), indicating that the additional variables in the more complex model contribute to a statistically significant improvement in modeling the pain outcome.

Overall, the results suggest that psychological factors like pain catastrophizing and mindfulness, as well as hormonal measures like cortisol levels, provide substantial additional information about postoperative pain beyond what is explained by demographic variables alone.
