---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(viridis)

expeditions <- tt_load("2020-09-22")$expeditions

# Process the data
expeditions_processed <- expeditions %>%
  filter(!is.na(peak_name), !is.na(season)) %>%
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != "Other") %>%
  group_by(peak_name, season) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(peak_name = fct_reorder(peak_name, n, .fun = sum))

# Create the plot
ggplot(expeditions_processed, aes(x = n, y = peak_name, fill = season)) +
  geom_col() +
  scale_fill_viridis_d() +
  theme_light() +
  labs(x = "Number of expeditions", y = "Peaks", fill = "season",
       title = "The 15 most popular peaks stacked by season of expedition") +
  theme(legend.position = "bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}

library(tidyverse)
library(tidytuesdayR)
library(scales)

# Load the data
tt <- tt_load("2019-02-19")
phd_by_field <- tt$phd_by_field

# Remove rows with NA in n_phds
phd_by_field <- phd_by_field %>%
  filter(!is.na(n_phds))

# Aggregate the data to count all phd-s by year and broad_field
phd_agg <- phd_by_field %>%
  group_by(year, broad_field) %>%
  summarise(total = sum(n_phds), .groups = "drop")

# Create the plot
ggplot(phd_agg, aes(x = year, y = total, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  labs(x = "", y= "", title = "Number of awarded Ph.D.-s in the US by year", color = "Broad field")
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
library(tidyverse)
library(tidytuesdayR)
library(scales)


tt <- tt_load("2019-11-05")
commute <- tt$commute
View(commute)

# Aggregate the data to count all commutes by state for each mode
commute_walk <- commute %>%
  filter(mode == "Walk") %>%
  group_by(state) %>%
  summarise(total_walk = sum(n), .groups = "drop")

commute_bike <- commute %>%
  filter(mode == "Bike") %>%
  group_by(state) %>%
  summarise(total_bike = sum(n), .groups = "drop")

# Merge the two data frames
commute_agg <- merge(commute_walk, commute_bike, by = "state")

# Add state region and abbreviation to the data frame
commute_agg <- merge(commute_agg, commute[, c("state", "state_region", "state_abb")], by = "state")

# Create the plot
ggplot(commute_agg, aes(x = total_walk, y = total_bike, color = state_region)) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abb), vjust = 0.5, hjust = 0.5, color ="black") +
  scale_x_log10(labels = comma_format()) +
  scale_y_log10(labels = comma_format()) +
  theme_light() +
  labs(x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", color = "State region", title = "Number of people walking vs. biking to work in each USA state")
```
