---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visualization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
# Load data
tt_data <- tt_load("2020-09-22")
expeditions <- tt_data$expeditions

# Remove unnecessary variables
expeditions <- expeditions %>% 
  select(peak_name, season)

# New variable for counting expeditions per each peak
expeditions <- expeditions %>% 
  group_by(peak_name) %>%
  mutate(expeditions_per_peak = n()) %>%
  arrange(desc(expeditions_per_peak))

# Get the top 15 targeted peaks
top_15_peaks <- expeditions %>%
  count(peak_name, sort = TRUE) %>%
  head(15)

# Filter the top 15 peaks
expeditions <- expeditions[expeditions$peak_name %in% top_15_peaks$peak_name, ]

# Prepare data for plotting
peak_season_counts <- expeditions %>%
  group_by(peak_name, season) %>%
  summarise(expeditions_count = n()) %>%
  ungroup() %>%
  group_by(peak_name) %>%
  mutate(total_expeditions = sum(expeditions_count)) %>%
  arrange(desc(total_expeditions))

# Create the barplot
ggplot(peak_season_counts, aes(x = expeditions_count, y = fct_reorder(peak_name, total_expeditions, .desc = FALSE), fill = season)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Number of Expeditions", y = NULL) +
  ggtitle("The 15 most popular peaks stacked by season of expedition") +
  scale_fill_viridis_d() +
  theme_light() +
  theme(legend.position = "bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
# Load data
phd <- tt_load("2019-02-19")
phd <- phd$phd_by_field

# Remove unnecessary variables
phd <- phd %>% 
  select(broad_field, year, n_phds)

# Aggregate data
## Remove NA-s
phd <- phd %>% 
  filter(!is.na(n_phds))

## Summarise n_phds per field per year
phd_summary <- phd %>% 
  group_by(broad_field, year) %>% 
  summarise(sum_phd = sum(n_phds))

# Plot data
ggplot(phd_summary, aes(x = year, y = sum_phd, color = broad_field)) +
  geom_line(linewidth = 1.2) +
  labs(x = NULL, y = NULL, color = "Broad Field") +
  ggtitle("Number of awarded Ph.D-s by research field and year") +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 5)) +
  theme_minimal() +
  scale_color_brewer(palette = "Dark2")
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
# Load data
commute <- tt_load("2019-11-05")
commute <- commute$commute

# Remove unnecessary variables
commute <- commute %>% 
  select(state_abb, mode, n, state_region)

# Aggregate data
commute <- commute %>%
  group_by(state_abb, mode) %>% 
  mutate(count = sum(n)) %>% 
  distinct(state_abb, mode, count, state_region)

# 2 columns for commuting mode for each state
commute_2 <- commute %>% 
  pivot_wider(names_from = mode, values_from = count)

# Plot data
ggplot(commute_2, aes(x = Walk, y = Bike, label = state_abb, color = state_region)) +
  geom_point(size = 2) +
  geom_text(color = "black") +
  labs(x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", color = "State Region") +
  ggtitle("Title number of people walking vs. biking to work in each USA state") +
  scale_color_discrete(name = "State Region") +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format(), breaks = c(1, 10, 100, 1000, 10000, 100000), expand = c(0, 1)) +
  theme_light()
```