---
title: "Assignment 6: Factor Analysis"
author: "Marton Kovacs / Zoltan Kekecs"
output: html_document
editor_options: 
  chunk_output_type: console
student: "Lili Kővári"
neptun ID: "YRUL7L"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("factoextra")
library(corrplot)
library("Hmisc")
library(tidyverse)
library(psych)
library(QuantPsyc)
library(energy)
library(factoextra)
```

# Introduction

In this lab assignment you will need to explore the factor structure of the Animal Rights Scale, a scale containing 28 items to measure attitudes towards animal experimentation and animal rights. Imagine that you are a researcher who is interested in the underlying factors that govern attitudes towards animal rights and the use of animals for different purposes. You have gathered data using the Animal Rights Scale (ARS) from 154 individuals in an online survey. Your goal is to explore the underlying factors.

# Dataset

You can load the dataset from the 'data/' folder.

The dataset includes the responses of 154 individuals on the following variables:

__ar1-ar28__ contain the data from the 28 items of the ARS. Participants had to rate their agreement with each statement separately on a 1-5 Likert scale with the following anchors: 1 - strongly disagree, 2 – disagree, 3 - no opinion, 4 – agree, 5 - strongly agree.

The questions in the ARS were the following:

  * __ar 1.__ Humans have no right to displace wild animals by converting wilderness areas into farmlands, cities, and other things designed for people.
  * __ar 2.__ Animal research cannot be justified and should be stopped.
  * __ar 3.__ It is morally wrong to drink milk and eat eggs.
  * __ar 4.__ A human has no right to use a horse as a means of transportation (riding) or entertainment (racing).
  * __ar 5.__ It is wrong to wear leather jackets and pants.
  * __ar 6.__ Most medical research done on animals is unnecessary and invalid.
  * __ar 7.__ I have seriously considered becoming a vegetarian in an effort to save animal lives.
  * __ar 8.__ Pet owners are responsible for preventing their pets from killing other animals, such as cats killing mice or snakes eating live mice.
  * __ar 9.__ We need more regulations governing the use of animals in research.
  * __ar 10.__ It is morally wrong to eat beef and other "red" meat.
  * __ar 11.__ Insect pests (mosquitoes, cockroaches, flies, etc.) should be safely removed from the house rather than killed.
  * __ar 12.__ Animals should be granted the same rights as humans.
  * __ar 13.__ It is wrong to wear leather belts and shoes.
  * __ar 14.__ I would rather see humans die or suffer from disease than to see animals used in research.
  * __ar 15.__ Having extended basic rights to minorities and women, it is now time to extend them also to animals.
  * __ar 16.__ God put animals on Earth for man to use. (invert)
  * __ar 17.__ There are plenty of viable alternatives to the use of animals in biomedical and behavioral research.
  * __ar 18.__ Research on animals has little or no bearing on problems confronting people. (invert)
  * __ar 19.__ New surgical procedures and experimental drugs should be tested on animals before they are used on people. (invert)
  * __ar 20.__ I am very concerned about pain and suffering in animals.
  * __ar 21.__ Since many important questions cannot be answered by doing experiments on people, we are left with no alternatives but to do animal research. (invert)
  * __ar 22.__ It is a violation of an animal's rights to be held captive as a pet by a human.
  * __ar 23.__ It is wrong to wear animal fur (such as mink coats).
  * __ar 24.__ It is appropriate for humans to kill animals that destroy human property, for example, rats, mice, and pigeons. (invert)
  * __ar 25.__ Most cosmetics research done on animals is unnecessary and invalid.
  * __ar 26.__ It is morally wrong to eat chicken and fish.
  * __ar 27.__ Most psychological research done on animals is unnecessary and invalid.
  * __ar 28.__ Hunters play an important role in regulating the size of deer populations. (invert)

You can get more information about the ARS here: http://core.ecu.edu/psyc/wuenschk/Animals/Anim-Rights-Q.htm

And also here: 

Wuensch, K. L., Jenkins, K. W., & Poteat, G. M. (2002). Misanthropy, idealism, and attitudes towards animals. _Anthrozoös, 15_, 139-149

Sharp, H. W., Wuensch, K. L., Eppler, M. A., & Harju, B. L. (2006, April). Narcissism, empathy, and attitudes towards animals. In _Spring Conference of the North Carolina Psychological Association and North Carolina Psychological Foundation, Charlotte, NC._

A few other questions were also included in the questionnaire:

__sex:__ The self reported sex of the participant. This is a categorical variable coded as 1 – female, 2 – male.

__party:__ Self reported party affiliation of the person (in the USA). This is a categorical variable coded as 1 - democrat, 2 - republican, 3 - other, 4 – none.

__liberal:__ This variable contains data from a question: please rate how conservative or liberal are you. On a scale of 1-5 where 1 means very conservative and 5 means very liberal. 

# Task

Your task is to do an exploratory factor analysis using the items in the ARS to identify the latent factors underlying the responses. First of all, start by exploring the descriptive statistics and correlations in the dataset to get more familiar with it and to identify any unusual cases or coding errors. Make sure to check the assumptions of factorability and multivariate normality and address them as necessary. You have a free hand in choosing the extraction and rotation methods. You can also exclude items if you see this necessary, but __do not exclude more than 8 items__ in this assignment. (If you still find the average communality below expectations, just report this as a limitation in your report, but continue the task). Keep notes of the steps and different setting/methods you tried during the exploratory factor analysis. 

_(The factor structure of this scale has been previously analyzed by others. If you want, you can use these previous research reports to guide your exploration, or you can ignore them. In any case, do not base your decisions solely on these research reports. Do your own work and base your decisions on your own findings on this dataset.)_

When you have arrived at the factor structure you consider final, give names to the factors you derived from the data. Save the factor scores and build a linear regression model to predict how conservative or liberal participants are (using the “liberal” variable as a dependent variable) with the factors you identified as the predictors.

__To simplify the task you can regard all likert scale variables (ar1-28 and liberal) as if they were continuous variables!__ So you do not have to use polychoric correlation for factor analysis and you do not have to perform ordinal regression.

# What to report

Report if you have found any unusual things (outliers or coding errors) in the dataset and how you dealt with them. Report the results of the assumption checks for factorability and multivariate normality. If any of the assumptions were found to be violated, report what was done to handle that. 

Report the number of factors you chose to keep in your final factor structure and give a rationale why. Include the parallel analysis scree plot in your report. Report the post-extraction eignevalues, variance explained, and cumulative variance explained by the final factors in a table format. Report the average post-extraction communality of the items. 

Report which rotation you chose to use (if any) and why. Report the final factor structure including the factor names. Also, report the post-extraction commonalities of each item and the loadings of the items on the final factors in a table format. (These can be reported in the same table). This table should contain the loadings that you used to interpret the factors in your analysis (e.g. the loadings listed in the rotated factor matrix or the pattern matrix). The table should be structured in a way to help the reader easily see which items load high on which factors.

Report if you have excluded any items, and give a rationale for each. 

Report which factor (if any) was the most influential predictor of how liberal a person is in the linear regression model and explain what do you base this assessment on.

# What to discuss

Talk about the limitations of your study and findings. 

# Solution

## Read the data

Read the Animal Rights Scale (ARQ) dataset from the 'data/' folder. Pay attention to the extension.

```{r}
data <- read.csv('C:/Users/Felhasználó/OneDrive/Asztali gép/assignments/3rdassignment/assignment_5_dataset.csv')
view(data) # assessing current dataset
```

## EDA

```{r}
str(data)

data %>%
  count(sex) # 124 women, 29 women participated, 1 did not share 

data %>%
  count(party) %>% # 30 dem, 34 rep, 9 other, 80 NA
  ggplot(aes(party,n))+
  geom_col() # most subjects did not provide information about their party affiliation as for those who did, seems like number of democrats and number of rebublicans is also the same, lets check the sex distribution between these two groups!

sex12 <- data %>%
  filter(party %in% c("1","2")) 
view(sex12) # we narrowed our further analyses, see how many subjects remained!

str(sex12) # 64 observations = 64 subjects. How about sexes?
sex12 %>%
  count(sex) # 45 women, 19 men. Does not seem equally distributed, right? Anyway, as a plot twist, see what's the outcome this way

sex12 %>%
  filter(party=="1") %>%
  count(sex) # 24 democrat women, 6 democrat men
sex12 %>%
  filter(party=="2") %>%
  count(sex) # 21 republican women, 13 democrat men

# how about liberality?
sex12 %>%
  count(liberal) %>% # 2 very conservative - 21 - 25 - 13 - 3 very liberal
  ggplot(aes(liberal,n))+
  geom_col() # seems like liberality is almost normally distributed

# here comes the fun part: the questionnaire! How about the scales? I noted that - at least in my opinion - there were items that are required to interpret in an inverted way (not to mention they are quite controversial); to make my work easier, I decided to eliminate these items from further analyses, these are variables 'ar16', 'ar19', 'ar21', 'ar24', 'ar28', I'd rather do it here, before other data manipulations

final <- sex12 %>%
  mutate(ar16=NULL,ar19=NULL,ar21=NULL,ar24=NULL,ar28=NULL)
view(final)
str(final) # from now on, we'll only work with 26 variables

means <- colMeans(final[ , 1:23]) # from 1 to 5, these are the means to each items
mean(means) # the absolute mean answer was 2,25, pretty "mean-ish" ranging form 1 to 5
sort(means) # so the lowest score was given to item3, indicating that people don't really wish to leave diary and egg out of their diet... the highest score was given to item25, showing people are pretty much against cosmetics research - well, who am I to disaggree?

# have a look at the individual means!
final$m <-apply(final[,1:23],1,mean)
view(final) # at this point this does not really make sense.. maybe some visualization helps:

final %>%
  ggplot(aes(liberal,m,color=sex, size=party))+
  geom_point() # probably not the best choice but somehow we can see a tendency... and that's what we want to test below
```

## Data manipulation

Recode the sex and party variables as factor type variables with the following levels:
  * sex: 1 - male, 2 - female
  * party: 1 - democrat, 2 - republican, 3 - other, 4 - none

```{r}
fin <- final %>%
  mutate(sex=recode(sex, '1'='female', '2'='male'),party=recode(party, '1'='democrat', '2'='republican'))
view(fin)

# note1: I wish to work only with democrats and republicans, therefore I only recoded these columns
# note2: according to the first explanation above, I used 1 as females, 2 as males :)
```

# Creating a correlation matrix

__Note:__ Remember to only include the variables of the questionnaire that will be part of the factor analysis.

```{r}
# I partialed out the variables above, however, it may be required to eliminate the row-wise mean column I created. Also, get rid of sex and party since they are categorical:
reduced <- fin %>%
  mutate(m=NULL,sex=NULL,party=NULL)
view(reduced)
```

Create the correlation matrix.

```{r}
cmat <- cor(reduced)
round(cmat, 2)
```

## Visualizing the correlation matrix

Create a visualization of the results of the correlation matrix.

```{r}
corrplot(cmat)
```

## Test for factorability

Calculate the KMO score.

```{r}
KMO(reduced) # KMO overall is > .60, so we can go along with these indicators
```

## Test for multivariate normality

```{r}
mvnorm.etest(reduced, R=100) # the Energy-test w/ 100 replicates showed a p value indicating that we dn't have evidence to presume our variables don't follow multivariate distribution
```

Test for skewness and kurtosis.

```{r}
mult.norm(reduced)$mult.test # another method, supporting the results of the preceding test
```

## Create scree plot

Create a scree plot to help the decision on how many factors to include.

```{r}
# first, we perform a principal component analysis (PCA)
princ <- prcomp(reduced, center = TRUE, scale. = TRUE)
princ

# then we calculate total variances
vars = princ$sdev^2 / sum(princ$sdev^2)
vars

# finally, we make the plot
qplot(c(1:24), vars) + 
  geom_line() + 
  xlab("dimension") + 
  ylab("eigenvalue")+
  ggtitle("Scree plot")

# see the explained variance percentages
percs <- print(vars)
sort(percs, decreasing=TRUE) # based on this I'm too afraid to exclude any. But according to Kaiser's rule, only those eigenvalues that exceed 1 should be kept. Let's see:
percs>0.01 # the first 19 components. Elbow method? The line flattens out from the 6th component which suggests we shall include the first 5 principal components
```

## Run the factor analysis

Run the factor analysis with the chosen number of factors.

```{r}
fa <- factanal(reduced, factors = 5, rotation = "varimax")
fa
```

Sort the communality scores in decreasing order.

```{r}
# we compute them by squaring loadings
communality <- sort(apply(fa$loadings^2, 1, sum),decreasing=TRUE)
communality
# we have some pretty high scores that can mirror an appropriate factor model. Another method:
comm2 <- 1 - apply(fa$loadings^2, 1, sum)
comm2

#plus: Another indicator might be: lower uniqueness scores:
uniq<- sort(fa$uniquenesses)
mean(uniq)
mean(communality) # almost equal but there is a slight difference in favor or the communality
```

Calculate the mean communality scores.

```{r}
meancom <- mean(communality)
meancom # 0.529
```

Show the factor loadings for the chosen factor structure.

```{r}
loadings <- fa$loadings
loadings
```

Visualize the factor structure.

```{r}
summary(loadings)

# to coincide columns:
format <- melt(loadings, measure=c("F1", "F2", "F3", "F4", "F5"), variable.name="Factor", value.name="Loading")

```

## Run linear regression

Calculate the factor scores.

```{r}
fscores <- factor.scores(reduced, fa)
fscores
scores<- fscores$scores
scores
```

Bind factor scores to the original dataset.

```{r}
orig <- cbind(reduced,scores)
view(orig)
```

Run the logistic regression.

```{r}

```
