---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(scales)

```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}

# Importing data

expeditions_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv")

view(expeditions_raw)

# Creating factors of the 15 most frequent peaks, and dropping the other factor

expeditions <- expeditions_raw %>%
  mutate(peak_name = fct_lump(peak_name, n = 15, other_level = "Other")) %>% 
  filter(peak_name != "Other") 

# Calculating the sum of expeditions

expeditions <- expeditions %>% 
  group_by(peak_name) %>% 
  mutate(sum_expeditions = n())

# Creating the plot

ggplot(expeditions, aes(y = fct_reorder(peak_name, sum_expeditions), fill = season)) + # Defining scales, the filling of bars, and ordering peaks, as the sum of expeditions
    geom_bar(show.legend = TRUE) + # Creating a bar plot, and showing the color palette
    scale_fill_viridis_d()+ # Selecting viridis palette
    labs(y = NULL, 
        x = "Number of expeditions",
        title = "The 15 most popular peaks stacked by season of expedition")+ # Adding labels and title  
    theme_light()+ # setting theme to theme light
    theme(legend.position = "bottom") # Setting legend to appear under the plot

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}

# Importing data

phd_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")

view(phd_raw)

# Counting all phds by broad fields

phd <- phd_raw %>%
  group_by(year, broad_field) %>% 
  mutate(sum_n_phds = sum(n_phds, na.rm = TRUE))

view(phd)

# Creating the plot

ggplot(phd, aes(x = year, y = sum_n_phds, color = broad_field)) + # Defining scales, adding colors indicating broad fields 
  geom_line(size = 1.2) + # Creating a line plot, and setting line size to 1.2
  scale_x_continuous(breaks = pretty_breaks()) + # Applying pretty_breaks() to x axis
  scale_y_continuous(labels = comma_format()) + # Setting labels of the y axis comma formatted
  labs(y = NULL, 
        x = NULL,
        title = "Number of awarded Ph. D.-s in the US by year")+ # Adding labels and title
  scale_color_brewer(palette = "Dark2",  name = "Broad field") + # Setting color palette to Dark2 and adding title to the color palette
  theme_minimal() # Setting theme to theme minimal




```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}

# Importing data

commute_raw <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")

# Counting all commutes by state and mode, adding state region to the new table as well

commute <- commute_raw %>%
  group_by(state_region, state, mode) %>%
  summarise(TotalCommutes = sum(n)) %>%
  spread(key = mode, value = TotalCommutes, fill = 0)

# Adding a column of state names abbrevations

commute <-  merge(commute, data.frame(state = state.name, abbrev = state.abb), by = "state")

view(commute)


ggplot(commute, aes(x = Walk, y = Bike, label = state, color = state_region)) + # Defining scales, adding colors indicating state region 
  geom_point(size= 2) + # Creating a scatter plot, setting point size to 2
  geom_text(aes(label = abbrev), color = "black", check_overlap = TRUE
) + # Adding labels with the abbreviations of state names, without labels, that would overlap
  labs(x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", title = "Number of people walking vs. biking to work in each USA state",  color = "State region") + # Adding labels and title 
  scale_x_log10(labels = scales::comma_format()) + # Applying log transformation, and comma formatting on the y-axis
  scale_y_log10(labels = scales::comma_format(), limits = c(25, NA)) + # Applying comma formatting on the x-axis, and setting the lower limit to 25
  theme_light() # Setting theme to theme light






```
