---
title: "Assignment 3: Model comparison"
author: "Marton Kovacs/Zoltan Kekecs"
output: html_document
editor_options: 
  chunk_output_type: console

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readxl)
library(psych) 
library(car)
library(lmtest)
library(lm.beta) 


```

In this lab assignment you are going to work with (simulated) data related to perioperative pain and its psychological and hormonal predictors. In the assignment you will assess the added benefit of including some psychological and hormonal predictors to the already established demographic predictors of pain.

In this assignment you will set up a hierarchical regression model to predict postoperative pain after wisdom tooth surgery. 

# Research problem

The amount of pain experienced around and after surgeries are highly variable between and within individuals. In order to improve surgical pain management regimens we need to understand what influences pain around surgical procedures and predict the amount of pain an individual will experience.

Your first study in this area is related to assessing the influence of trait and state psychological measures on pain, and to see whether taking into account these variables can improve our understanding of postoperative pain.

# Procedures and measures

Use the data file called ‘assignment_3_dataset’, from the 'data/' folder.

You have collected data from 160 adults who were scheduled to undergo surgical extraction of the third mandibular molar (wisdom tooth surgery). Patients filled out a form in the waiting room before their surgery. The form contained questions about their sex, age, and weight, and psychological questionnaires assessing anxiety, pain catastrophizing, and mindfulness (see descriptions below). You also got blood samples and saliva samples from participants in the waiting room 5 minutes before their operations to determine the mark (a component of the blood) and salivary cortisol levels of participants. Participants were contacted 5 hours after the surgery to see how much pain they were experiencing. The __level of pain__ at that moment was recorded using a numerical rating scale using a __scale of 0 to 10__, where 0 means “no pain” and 10 means “worst pain I can imagine”. 

__The State Trait Anxiety Inventory:__ T measures trait anxiety on a scale of 20 to 80, higher scores mean higher anxiety. Anxiety has been found in many studies to positively correlate with the level of pain experienced. This is __variable STAI_trait__ in the dataset.

__The Pain Catastrophizing Scale__ measures the extent of pain catastrophizing, which is characterized by a tendency to magnify the threat value of a pain stimulus and to feel helpless in the presence of pain, as well as by a relative inability to prevent or inhibit pain-related thoughts in anticipation of, during, or following a painful event. The total score on this scale ranges from 0 to 52, higher scores mean higher catastrophizing. Pain catastrophizing is one of the well-established predictors of clinical pain. This is __variable pain_cat__ in the dataset.

__The Mindful Attention Awareness Scale (MAAS)__ measures dispositional mindfulness, which may be described as a tendency to turn attention to present-moment experiences in an open, non-judgmental way. The MAAS total score ranges from 1 to 6 (an average of the item scores), with higher scores representing higher dispositional mindfulness. Trait mindfulness has been theorized to serve as a protective factor against pain, as the individual would be more objective about their pain experience and tend to associate less discomfort, despair, and hopelessness to the pain-related sensations. This is __variable mindfulness__ in the dataset.

__Cortisol__ is a stress hormone associated with acute and chronic stress. Cortisol levels are thought to be positively associated with pain experience. Cortisol can be __measured from both blood and the saliva__, although, mark cortisol is often regarded in medical research as more reliably related to stress (serum is a component of the blood plasma). These are __variables cortisol_serum__, and __cortisol_saliva__ in the dataset.

# Research question

Previous studies and meta-analyses showed that age and sex are often predictors of pain (age is negatively associated with pain, while sex is a predictor more dependent on the type of the procedure). You would like to determine the extent to which taking into account psychological and hormonal variables aside from the already used demographic variables would improve our understanding of postoperative pain.

To answer this research question you will __need to compare two models__ (with a hierarchical regression). The __simpler model__ should contain __age and sex as predictors of pain__, while the __more complex model__ should contain the __predictors: age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures__. Notice that the predictors used in the simpler model are a subset of the predictors used in more complex model. __You will have to do model comparison to assess whether substantial new information was gained about pain in the more complex model compared to the simpler model.__  

# What to report

As usual, before you can interpret your model, you will need to run data and model diagnostics. First, check the variables included in the more complex model (age, sex, STAI, pain catastrophizing, mindfulness, and cortisol measures as predictors, and pain as an outcome) for __coding errors__, and the model itself for __influential outliers__ (for example using Cook’s distance). Furthermore, check the final model to see if the __assumptions of linear regression hold true__, that is, __normality__ (of the residuals), __linearity__ (of the relationship), __homogeneity of variance__ (also called homoscedasticity) and that there is no excess __multicollinearity__ (“uncorrelated predictors” in Navarro’s words). If you find anything amiss during these checks, make the appropriate decision or correction and report your findings and actions in your report. 

__Note:__ If you do any changes, such as exclude cases, or exclude predictors from the model, you will have to re-run the above checks for your final data and model.

Report the results of the simpler model and the more complex model. For both models you should report the model test statistics (adj.R2, F, df, and p value). Also, report the statistics describing the coefficients of the predictors in a table format (unstandardized regression coefficients and 95% confidence intervals, standardized regression coefficients (B and Beta values), and p values).

Write up the regression equation of the more complex model in the form of 𝑌 = 𝑏0 + 𝑏1 ∗ X1 + 𝑏2 ∗ X2 +…+ bn * Xn, in which you use the actual regression coefficients of your models. (b0 stands for the intercept and b1, b2 … bn stand for the model coefficients for each of the predictors, and X1, X2, … Xn denote the predictors).

Compare the two models in terms of how much variance they explain of pain’s variability in the sample. Report Akaike information criterion (AIC) for both models and the F test statistic and p value of the likelihood ratio test comparing the two models.

# What to discuss

In your discussion of the findings, briefly interpret the results of the above analyses, and indicate whether you think that anything was gained by including the psychological and hormone measures in the model.

# Solution

## Read the data

Read the dataset used in this assignment. Pay attention to the extension of the datafile.

```{r}

pain_raw <- read_excel("data\\assignment_3_dataset.xlsx")

view(pain_raw)


```

## Data and model diagnostics 
### Data diagnostics
#### Descriptives of the variables

Run an exploratory data analysis (EDA) to investigate the dataset.

```{r}

# Checking variable: pain

ggplot(pain_raw, aes(pain))+
  geom_bar()

describe(pain_raw$age)

pain_raw %>% 
  filter(is.na(age) == TRUE) %>% 
  count()

describe(pain_raw$pain)

# I found a coding error in the variable: 50 is out of the range of the variable, probably a typo and should be five

# Checking variable: age

ggplot(pain_raw, aes(age))+
  geom_histogram()

describe(pain_raw$age)

pain_raw %>% 
  filter(is.na(age) == TRUE) %>% 
  count()


# Checking variable: sex

pain_raw %>% 
  filter(is.na(sex) == TRUE) %>% 
  count()

ggplot(pain_raw, aes(sex))+
  geom_bar()

# I found a coding error, in one cell woman is written, while in others the term female is used


# Checking variable: STAI_trait

pain_raw %>% 
  filter(is.na(STAI_trait) == TRUE) %>% 
  count()

ggplot(pain_raw, aes(STAI_trait))+
  geom_bar()

describe(pain_raw$STAI_trait)

# Checking variable: pain_cat

pain_raw %>% 
  filter(is.na(pain_cat) == TRUE) %>% 
  count()

ggplot(pain_raw, aes(pain_cat))+
  geom_bar()

describe(pain_raw$pain_cat)

# Checking variable: mindfulness

pain_raw %>% 
  filter(is.na(mindfulness) == TRUE) %>% 
  count()

ggplot(pain_raw, aes(mindfulness))+
  geom_bar()

describe(pain_raw$mindfulness)

# I found a coding error, the value 6.22 is out of range, this value should be removed, since it is difficult to determine what the real value might have been

# Checking variable: cortisol_serum

pain_raw %>% 
  filter(is.na(cortisol_serum) == TRUE) %>% 
  count()

ggplot(pain_raw, aes(cortisol_serum))+
  geom_bar()

describe(pain_raw$cortisol_serum)

# Checking variable: cortisol_saliva

pain_raw %>% 
  filter(is.na(cortisol_saliva) == TRUE) %>% 
  count()

ggplot(pain_raw, aes(cortisol_saliva))+
  geom_bar()

describe(pain_raw$cortisol_saliva)

```

#### Correct coding errors

If you find values in the dataset during the EDA, that are not correct based on the provided descriptions of the variables of the dataset please correct them here.

```{r}

# Correcting coding error in pain: replacing 50 with 5 

pain <- pain_raw %>% 
  mutate(pain = replace(pain, pain == 50, 5))

ggplot(pain, aes(pain))+
  geom_bar()

# Correcting coding error in sex: replacing woman with female

pain <- pain %>% 
  mutate(sex = str_replace(sex, "woman", "female"))

ggplot(pain, aes(sex))+
  geom_bar()

# Correcting coding error in mindefullnes: replacing 6.22 with "NA"

pain <- pain %>% 
  mutate(mindfulness = replace(mindfulness, mindfulness == 6.22, NA))

pain %>% 
  filter(is.na(mindfulness) == TRUE) %>% 
  count()

ggplot(pain, aes(mindfulness))+
  geom_bar()

describe(pain$mindfulness)

```

### Model diagnostics
#### Build the more complex model

In order to test the more complex model for outliers and to test the assumptions first build the model.

```{r}

mdl_complex <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_serum + cortisol_saliva, data = pain)


```

#### Checking for influential outliers

Check for outlier values in the model.

```{r}

# Calculating Cook's distance

cooksd <- cooks.distance(mdl_complex)

# Plotting Cook's distance

mdl_complex %>%
  plot(which = 4)

# Calculating which datapoints, have a Cook's distance greater than 4/N

influential_obs <- which(cooksd > 4 / nrow(pain))

# Marking them on the plot

text(influential_obs, cooksd[influential_obs], labels = influential_obs, col = "red", cex = 0.8, pos = 4)

# There are some data points, which have a greater Cook's distance than 4/N, but neither of them is greater than 1. I decided to check the assumptions, before making a decision about them

```

#### Checking assumptions

Check the normality assumption.

```{r}

# Creating QQ plot

mdl_complex %>%
  plot(which = 2)

# The distribution of the data points is not very different from the line marking normal distribution

# Creating histogram of residuals

residuals_mdl_complex = enframe(residuals(mdl_complex))

residuals_mdl_complex %>%
  ggplot() +
  aes(x = value) +
  geom_histogram()

describe(residuals(mdl_complex))

# The skew and kurtosis of the distribution is smaller than one, which means, that the normality of the distribution is not violated



```

Check the linearity assumption.

```{r}

mdl_complex %>%
  residualPlots()

# The plots are somewhat curved, but none of the tests are significant, so our model may meet the assumption of linearity

```

Check the homoscedasticty assumption (homogeneity of variance).

```{r}

# Creating a plot to check homoscedasticty

mdl_complex %>%
  plot(which = 3)

# NCV test

mdl_complex %>%
  ncvTest() 

# Breush-Pagan test

mdl_complex %>%
  bptest() 

# Neither of the tests are significant, so the assumption of homoscedasticty is not violated

```

Check the multicollinearity assumption.

(VIF above 5), or a VIF threshold of 3 is recommended in this paper: http://onlinelibrary.wiley.com/doi/10.1111/j.2041-210X.2009.00001.x/full

Some info about VIF: 
https://statisticalhorizons.com/multicollinearity
http://blog.minitab.com/blog/understanding-statistics/handling-multicollinearity-in-regression-analysis

```{r}

# Calculating VIF values

mdl_complex %>%
  vif()

# The VIF value of the cortisol measures (cortisol_mari and cortisol_saliva) is greater than 5

```

### Making decision based on model diagnostics

If based on the assumption tests you decide to drop a predictor variable you should do that here. Create your updated model.

```{r}

# Creating a plot of all variables included in the modell, to see how they correlate with each other

pain %>%
  select(pain, age, sex, STAI_trait, pain_cat, mindfulness, cortisol_serum, cortisol_saliva) %>%
  pairs.panels(col = "red", lm = T)

# The two cortisol measures have a high correlation (0.91), suggesting, one of them should be excluded from the model. I decided to exclude the variable cortisol_serum, since it is more difficult to collect blood samples from participants than saliva samples, so the replication of the study would be easier.

mdl_complex2 <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva, data = pain)

```

#### Checking outliers of the updated model

```{r}

# Calculating Cook's distance

cooksd <- cooks.distance(mdl_complex2)

# Plotting Cook's distance

mdl_complex2 %>%
plot(which = 4)

# Calculating which datapoints, have a Cook's distance greater than 4/N

influential_obs2 <- which(cooksd > 4 / nrow(pain))

#  Marking them on the plot

text(influential_obs, cooksd[influential_obs2], labels = influential_obs2, col = "red", cex = 0.8, pos = 4)

# There are some data points, which have a greater Cook's distance than 4/N, but neither of them is greater than 1. I decided to check the assumptions, before making a decision about them


```

#### Checking assumptions of the updated model

Normality assumption

```{r}

# Creating QQ plot

mdl_complex2 %>%
  plot(which = 2)

# # The distribution of the data points is not very different from the line marking normal distribution

# Creating histogram of residuals

residuals_mdl_complex2 = enframe(residuals(mdl_complex2))

residuals_mdl_complex2 %>%
  ggplot() +
  aes(x = value) +
  geom_histogram()

describe(residuals(mdl_complex2))

# The skew and kurtosis of the distribution is smaller than one, which means, that the normality of the distribution is not violated




```

Linearity assumption

```{r}

mdl_complex2 %>%
  residualPlots()

# The plots are somewhat curved, but none of the tests are significant, so our model may meet the assumption of linearity

```

Homoscedasticty assumption (homogeneity of variance)

```{r}

# Creating a plot to check homoscedasticty

mdl_complex2 %>%
  plot(which = 3)

# NCV test

mdl_complex2 %>%
  ncvTest() 

# Breush-Pagan test

mdl_complex2 %>%
  bptest() 

# Neither of the tests are significant, so the assumption of homoscedasticty is not violated


```

Multicollinearity assumption

```{r}

# Calculating VIF values

mdl_complex2 %>%
  vif()

# Neither VIF value is grater than 3, so the assumption of multicollinearity is not violated


```

## Model comparison

Create the simple model and get the results of the model that needs to be reported based on the What to report section.

```{r}

pain <- na.omit(pain[c("pain", "age", "sex", "STAI_trait", "pain_cat", "mindfulness", "cortisol_saliva")])

# Since one value of mindfulness was omitted, this row of data shoul be dropped from the dataset of the simple model as well

# Creating the simple model

mdl_simple <- lm(pain ~ age + sex, data = pain)

```

Create the more complex model based on the results of the model diagnostics. Also, get the results that needs to be reported based on the What to report section.

```{r}

mdl_complex_final <- lm(pain ~ age + sex + STAI_trait + pain_cat + mindfulness + cortisol_saliva, data = pain)


```

Compare the two models.

```{r}

summary(mdl_complex_final)

AIC(mdl_complex_final)

confint(mdl_complex_final)

lm.beta(mdl_complex_final)

summary(mdl_simple)

AIC(mdl_simple)

confint(mdl_simple)

lm.beta(mdl_simple)

# Coef_table function (source: https://github.com/kekecsz/PSZB17-210-Data-analysis-seminar/blob/master/seminar_07/S07_Linearis%20regresszio.pdf)

coef_table = function(model){
  require(lm.beta)
  mod_sum = summary(model)
  mod_sum_p_values = as.character(round(mod_sum$coefficients[,4], 3))
  mod_sum_p_values[mod_sum_p_values != "0" & mod_sum_p_values != "1"] = substr(mod_sum_p_values[mod_sum_p_values != "0" & mod_sum_p_values != "1"], 2, nchar(mod_sum_p_values[mod_sum_p_values != "0" & mod_sum_p_values != "1"]))
  mod_sum_p_values[mod_sum_p_values == "0"] = "<.001"
  mod_sum_table = cbind(as.data.frame(round(cbind(coef(model), confint(model), c(0, lm.beta(model)$standardized.coefficients[c(2:length(model$coefficients))])), 2)), mod_sum_p_values)
  names(mod_sum_table) = c("b", "95%CI lb", "95%CI ub", "Std.Beta", "p-value")
  mod_sum_table["(Intercept)","Std.Beta"] = "0"
  return(mod_sum_table)
}

anova(mdl_simple, mdl_complex_final)

```

As the complex model I built a linear regression model in which I predicted the the amount of pain experienced around and after surgeries with the participants' age, sex, score on The State Trait Anxiety Inventory, Pain Catastrophizing Scale, The Mindful Attention Awareness Scale and their saliva sample cortisol level. 

A multiple regression model was  better at predicting the experienced pain than the null model. The model yielded 31% of the variance of the housing (F (7, 151) = 11.58, p < 0.001, Adj. Rˆ2 = 0.31, AIC = 530.06"

```{r}

# Creating a table with the statistics describing the coefficients

coef_table(mdl_complex_final)

```

As the simple model I built a linear regression model in which I predicted the the amount of pain experienced around and after surgeries with the participants' age and sex. 

A multiple regression model was  better at predicting the experienced pain than the null model. The model yielded 6% of the variance of the housing (F (2, 156) = 6.815, p = 0.003, Adj. Rˆ2 = 0.06, AIC = 575.06"

```{r}

# Creating a table with the statistics describing the coefficients

coef_table(mdl_simple)

```

Regression equation of the more complex model

$Y = 1.29 + (-0.2) * X1 + 0.32 * X2 + 0.08 * X3 + (-0.03) * X4 + 0.10 * X5 + (-0.12) * X6 + 0.59 * X7$



The model containing the psychological and hormonal predictors explained more variance (31%) than the model containing only the sex and age of participants (6%). The Akaike information criterion for the complex model is 530.06 and 575.06 for the simple model, suggesting that the complex model fits the data significantly better. The ANOVA results comparing the two models indicate a significant difference in model fit [F(5, 151) = 12.479, p < 0.001]. The inclusion of additional predictors in the complex model  significantly (ΔRSS = 96.33, Δdf = 5, p < 0.001) reduced the residual sum of squares (RSS) compared to Model 1 (ΔRSS = 96.33, Δdf = 5, p < 0.001), suggesting that the expanded model provides a better fit to the data.


Discussion

The results suggest, that the hormonal and psychological predictors added to the model resulted in a significantly better fit, suggesting that that trait anxiety,tendency for pain catastrophizing, dispositional mindfulness and cortisol level indicating acute and chronic stress affects the pain experienced postoperative pain.









