---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(scales)
tt1.data <- tt_load("2020-09-22")
tt2.data <- tt_load("2019-02-19")
tt3.data <- tt_load("2019-11-05")
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- tt1.data$expeditions

expeditions <- expeditions %>%
  mutate(peak = fct_lump(peak_name, n = 15, other_level = "Other"),
         season = as.factor(season))

expeditions.season <- expeditions %>%
  group_by(peak, season) %>%
  summarise(Count = n()) %>%
  filter(!is.na(peak) & peak != "Other")


ggplot(expeditions.season, aes(x = Count, y = fct_reorder(peak, Count, .fun = sum), fill = season)) +
  geom_bar(position ="stack", stat = "identity") +
  scale_fill_viridis_d() +
  labs(y = "", x = "Number of expeditions") +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = "The 15 most popular peaks stacked by season of expedition")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd.data.raw <- tt2.data$phd_by_field

phd.data <- phd.data.raw %>%
  replace(is.na(.), 0) %>%
  group_by(broad_field, year) %>%
  summarise(Count = sum(n_phds))

ggplot(phd.data, aes(year, Count, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = pretty_breaks(n = 5)) +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "Number of awarded Ph.D.-s in the Us by year", color = "Broad field") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute.data.raw <-tt3.data$commute

commute.data <- commute.data.raw %>%
  group_by(state_abb, mode, state_region) %>%
  summarise(count = sum(n)) %>%
  pivot_wider(names_from = mode,
              values_from = count) %>%
  filter(!is.na(state_abb))

ggplot(commute.data, aes(Walk, Bike, color = state_region, label = state_abb)) +
  geom_point(size = 2) +
  geom_text(color = "black") +
  labs(title = "Title number of people walking vs. biking to work in each USA state", x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", color = "State region") +
  scale_x_log10(labels = comma_format(), ) +
  scale_y_log10(labels = comma_format(),
                limits = c(25, 150000)) + 
  theme_light()

  
```
