---
title: 'Assignment 2: Data visualization'
author: "Cintia Anna Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visualization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(ggplot2)
library(readr)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv')
```

```{r}
expeditions_filtered <- expeditions %>%
  mutate(peak_top15 = forcats::fct_lump(peak_name, n = 15)) %>%
  filter(peak_top15 != "Other")
```

```{r}
ggplot(expeditions_filtered, aes(x = fct_reorder(peak_name, desc(expedition_id), length), fill = season)) +
  geom_bar(position = "stack") +
  coord_flip() +
  labs(y = "Number of expeditions",
       title = "The 15 most popular peaks stacked by season of expedition") +
  theme_light() +
  theme(axis.title.y = element_blank(),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.box = "horizontal") +
  scale_fill_viridis_d() 
```


## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_by_field <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")
```

```{r}
phd_aggregated <- phd_by_field %>%
  group_by(broad_field, year) %>% 
  summarise(phd_count = sum(n_phds, na.rm = TRUE))
```

```{r}
ggplot(phd_aggregated, aes(x = year, y = phd_count, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  labs(title = "Numer of awarded Ph.D.-s in the US by year",
       color = "Broad field") +
  theme_minimal() +
  theme(axis.title.x = element_blank(),
        axis.title.y = element_blank())
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")
```

```{r}
commute_aggregated <- commute %>%
  group_by(state_region, state, mode) %>% 
  summarise(mode_count = sum(n, na.rm = TRUE)) %>%
  spread(key = mode, value = mode_count, fill = 0)

state_abbrev <- setNames(state.abb, state.name)
```

```{r}
ggplot(commute_aggregated, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  geom_text(aes(label = state_abbrev[state]), colour = "black") +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format()) +
  labs(x = "Number of ppl walking to wrok (log N)",
       y = "Number of ppl biking to wrok (log N)",
       title = "Number of people walking vs. biking to work in each USA state",
       color = "State region") +
  theme_light()
```