---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
url <- 'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv'

expeditions <- readr::read_csv(url) 

str(expeditions)

library(dplyr)
library(ggplot2)
library(forcats)
install.packages("viridis")
library(viridis)

top_peaks <- expeditions %>%
  count(peak_name) %>%
  top_n(15) %>%
  arrange(desc(n)) %>%
  pull(peak_name)
print(top_peaks)

season_counts <- filtered_data %>%
  filter(peak_name %in% top_peaks) %>%
  count(peak_name, season) %>%
  arrange(peak_name, desc(n))

print(season_counts, n = Inf)


ggplot(season_counts, aes(x = n, y = fct_reorder(peak_name, n, .fun = sum), fill = season)) +
  geom_bar(stat = "identity") +
  labs(x = "Number of expeditions", y = "", title = "The 15 most popular peaks stacked by season of expedition") +
  scale_fill_viridis_d() +
  theme_light() +
  theme(axis.text.y = element_text(hjust = 0.5),
        legend.position = "bottom")

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}

url2 <- 'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv'
phd_by_field <- readr::read_csv(url2)
str(phd_by_field)
head(phd_by_field)

phd_agg <- phd_by_field %>%
  filter(!is.na(n_phds)) %>%
  group_by(year, broad_field) %>%
  summarise(total_phds = sum(n_phds), .groups = "drop")
print(phd_agg)

library(scales)

ggplot(phd_agg, aes(x = year, y = total_phds, color = broad_field)) +
  geom_line(size = 1.2) +
  labs(x = "", y = "", color = "Broad Field", title="Number of awarded Ph.D.-s in the US by year") +
  theme_minimal() +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_color_brewer(palette = "Dark2")

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}

url3 <- 'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv'
commute <- readr::read_csv(url3)
head(commute)
str(commute)

allamonkent <- commute %>%
  group_by(state_abb, state_region) %>%
  summarise(gyalogolas = sum(n[mode == "Walk"]),
            biciklizes = sum(n[mode == "Bike"]))
print(allamonkent)

ggplot(allamonkent, aes(x = gyalogolas, y = biciklizes, color = state_region)) +
  geom_point(size=2) +
  labs(x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", color = "State region", title = "Number of people walking vs biking to work in each USA state") +
  theme_light() +
  geom_text(aes(label = state_abb), vjust = 1, hjust = -0.2) +
  scale_x_log10(labels = comma_format()) +
  scale_y_log10(labels = comma_format()) 

```
